---
title: "Foodborne Illnesses" ##TODO
author: Amit, Devon, Kennedy
date: 2020-11-27
output: html_notebook
---

```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(usmap)
library(modelr)
library(splines)
options(na.action = na.warn)
```

```{r setup, echo = FALSE, message = FALSE}
df <- read_csv("data_outbreaks.csv")
df %>%
  filter(Species!="NA",
         Hospitalizations!="NA",
         State!="NA",
         Fatalities!="NA",
         Location!="NA",
         Food!="NA") -> df.filtered
```
Heatmap of Years and Months (now from Jan->Dec, and with custom colours)
```{r heatmap_years_months, echo = FALSE}
df %>% mutate(month_dec = match(Month, month.name)) %>%
  count(month_dec, Year) %>% ggplot(mapping = aes(x = month_dec, y= Year)) +
  geom_tile(mapping = aes(fill =n), na.rm = FALSE)+
  theme_bw()+theme_minimal() +
  labs(title="Cases month wise from 1998-2015", x="Month", y="Year")+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line()) +
  scale_x_continuous("Months", 1:12, labels = month.abb) +
  ### Deciding on either month.name or month.abb (month.name currently too squished together to read
  scale_fill_gradient(low="white", high="red") +
  labs(fill = "Number of Cases")
```
Number of Illnesses per State
```{r ill_per_state, echo = FALSE}
df %>%
  mutate(fips = fips(State)) %>%
  group_by(fips,State) %>%
  summarise(n=sum(Illnesses), .groups = "keep") %>%
  plot_usmap(data = .,values = "n", color = "black") +
  scale_fill_continuous(low = "white",
                        high = "red",
                        name = "Number of Illnesses",
                        label = scales::comma
                        ) +
  theme(legend.position = "right") +
  labs(title = "Illnesses per State")
```

```{r location_vs_illnesses, echo = FALSE, message = FALSE, include = FALSE}
##OLD PLOT CLEANED UP
df %>%
  filter(Location!="NA",Location!="Unknown") %>%
  group_by(Location) %>%
  summarise(illness_location=sum(Illnesses), .groups = "drop") %>%
  arrange(desc(illness_location)) %>%
  filter(illness_location>2000)%>%ggplot()+
  geom_bar(mapping = aes(y=Location, x=log2(illness_location)),
              fill="red", stat = 'identity')+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))
```
Number of Illnesses at each Location Type <br />
Could use some more work <br />
I prefer this version over the old version because it doesn't exclude values under 2000, but also doesn't have two locations in one value.
```{r locations_vs_illnesses_v2, echo = FALSE}
separate_rows(df, Location, sep = "; ") %>%
  filter(!is.na(Location),Location != "Unknown", Location != "Other(Descripe In Remarks)") %>%
  group_by(Location) %>%
  summarise(illness_location=sum(Illnesses), .groups = "drop") %>%
  arrange(illness_location) %>%
  ggplot() +
  geom_bar(mapping = aes(y=Location, x=log2(illness_location)),
           fill="red",
           stat = 'identity') +
  xlab(bquote(~Log[2]~ "of Illness Location")) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))
```
Species responsible for highest number of cases <br />
I removed the legend because it didn't display any new information, and cleaned up the code a bit. This plot probably still needs some work.
```{r species_number, echo = FALSE}
separate_rows(df, Species, sep = "; ") %>%
  filter(Species!="NA") %>%
  group_by(Species) %>%
  summarise(ill_species =sum(Illnesses), .groups = "drop") %>%
  arrange(ill_species) %>%
  filter(ill_species > 64) %>%
  ggplot() +
    geom_col(mapping = aes(y=Species,x=log2(ill_species), fill=Species)) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black"),
          legend.position = "none") +
  ggtitle("Number of Illnesses Caused by Species (LOG2)") +
  xlab(bquote(~Log[2]~ "of Number of Illnesses")) +
  ylab("Name of Species")

```
I tidied up a lot of this code, but I couldn't replicate the final model, so I left it around. <br />
I also created a
```{r plot, echo = FALSE, message = FALSE}
#Species responsible for highest number of hospitalizations

separate_rows(df, Species, sep = "; ") %>%
  filter(!is.na(Species), !is.na(Hospitalizations), !is.na(Fatalities), !is.na(Illnesses)) %>%
  group_by(Species) %>%
  summarise(ill_species =sum(Illnesses),
            hospitalizations=sum(Hospitalizations),
            fatalities=sum(Fatalities),
            .groups = "drop") %>%
  filter(ill_species < 40000, hospitalizations < 600) -> df.species
ggplot(df.species) +
  geom_point(position = "jitter", aes(x= hospitalizations, y= ill_species)) +
  ggtitle("Number of Illnesses and Hospitalizations Caused by Species") +
  xlab("Number of Hospitalizations") +
  ylab("Number of Illnesses")

par <- coef(lm(ill_species~hospitalizations, data = df.species))

df.species %>%
  ggplot(mapping=aes(y=ill_species,x=hospitalizations)) +
  geom_point(position = "jitter") +
  geom_abline(intercept = par[1], slope = par[2]) +
  ggtitle("Linear Model of Number of Illnesses and Hospitalizations Caused by Species") +
  xlab("Number of Hospitalizations") +
  ylab("Number of Illnesses")

model <- lm(ill_species~ns(hospitalizations,3), data = df.species)
grid <- df.species %>% data_grid(hospitalizations) %>%
  add_predictions(model)

df.species %>%
  ggplot(aes(hospitalizations)) +
  geom_point(aes(y=ill_species),position = "jitter") +
  geom_line(aes(y=pred), data=grid, colour="red", size=1) +
  ggtitle("Model of Number of Illnesses and Hospitalizations Caused by Species") +
  xlab("Number of Hospitalizations") +
  ylab("Number of Illnesses")
```
Modeling from logs shows a clearer correlation, but the data is transformed to account for log(0).
```{r log_model, echo = FALSE, message = FALSE}
separate_rows(df, Species, sep = "; ") %>%
  filter(!is.na(Species), !is.na(Hospitalizations), !is.na(Fatalities), !is.na(Illnesses)) %>%
  group_by(Species) %>%
  summarise(ill_species =sum(Illnesses),
            hospitalizations=sum(Hospitalizations),
            fatalities=sum(Fatalities),
            .groups = "drop") %>%
  #filter(hospitalizations > 0, ill_species > 0) %>%   ## The other option instead of adding 1.
  mutate(hospitalizations = log2(hospitalizations + 1),
         ill_species = log2(ill_species + 1)) -> df.species_l

model <- lm(ill_species~ns(hospitalizations,3), data = df.species_l)
grid <- df.species_l %>% data_grid(hospitalizations) %>%
  add_predictions(model)
df.species_l %>%
  ggplot(aes(hospitalizations)) +
  geom_point(aes(y=ill_species),position = "jitter") +
  geom_line(aes(y=pred), data=grid, colour="red", size=1) +
  ggtitle("Model of Number of Illnesses and Hospitalizations Caused by Species (LOG2)") +
  xlab(bquote(~Log[2]~ "of Hospitalizations")) +
  ylab(bquote(~Log[2]~ "of the Number of Illnesses"))

```
This is the old plot, it appears more smooth than the one I recreated.
```{r old_plot, echo = FALSE, message = FALSE}
## I restructured the first few variables so that they wouldn't output to the md file.
df %>%
  filter(Species!="NA") %>%
  group_by(Species) %>%
  summarise(ill_species =sum(Illnesses), .groups = "drop") %>%
  arrange(ill_species) -> species_cases

df %>%
  filter(Species!="NA",Hospitalizations!="NA") %>%
  group_by(Species) %>%
  summarise(hospitalizations=sum(Hospitalizations), .groups = "drop") %>%
  arrange(desc(hospitalizations)) -> species_hospitalization

#Species associated with the highest number of fatalities
df %>%
  filter(Species!="NA",Fatalities!="NA") %>%
  group_by(Species) %>%
  summarise(fatalities=sum(Fatalities), .groups = "drop") %>%
  arrange(desc(fatalities)) -> species_fatalities

#Combining both species_cases,species_hospitalizations, and species_fatalities
species_1 <- merge(species_hospitalization, species_fatalities)

species_merge <- merge(species_1,species_cases)

#Scatter plot and model between hospitalizations and illnesses
species_subMerge <- species_merge%>%filter(hospitalizations<600,
                                   ill_species<40000)

model <- lm(ill_species~ns(hospitalizations,3), data = species_subMerge)
grid <- species_subMerge %>% data_grid(hospitalizations) %>%
  add_predictions(model)

df.species%>%ggplot(aes(hospitalizations))+
  geom_point(aes(y=ill_species),position = "jitter")+
  geom_line(aes(y=pred), data=grid, colour="red", size=1)+
  ggtitle("Model between illness and hospitalizations")
```
for some reason the scatter plots don't want to plot in with the other things
```{r, echo = FALSE, include = FALSE}
ggplot(data = df.species) +
  geom_point(aes(x = hospitalizations, y = ill_species))
ggplot(data = df.species) +
  geom_point(aes(x = fatalities, y = ill_species))
ggplot(data = df.species) +
  geom_point(aes(x = hospitalizations, y = fatalities))
```

```{r more_stuff, echo = FALSE}
# scatter plot in columnwise
par(mfrow=c(2,2))
plot(df.species$hospitalizations, df.species$ill_species)
plot(df.species$fatalities, df.species$ill_species)
plot(df.species$hospitalizations, df.species$fatalities)
```

```{r even more stuff, echo = FALSE}
#highest hospitalized cases across all states
df %>%
  filter(Hospitalizations>0) %>%
  group_by(State) %>%
  summarise(Total_hospitalized_cases=sum(Hospitalizations), Total_illness=sum(Illnesses), .groups = "drop") %>%
  arrange(desc(Total_hospitalized_cases), desc(Total_illness)) -> Highest_hospitalized_cases

#Relation between Illnesses and hospitalizations
ggplot(Highest_hospitalized_cases) +
  geom_freqpoly(mapping = aes(x=Total_hospitalized_cases, color="green") )

#Number of illnesses per year in descending order
df.filtered %>%
  group_by(Year,Location, Status) %>%
  mutate(Hospitalizations = replace_na(Hospitalizations,0)) %>%
  mutate(Fatalities = replace_na(Fatalities,0)) %>%
  summarise(outbreaks = n(),
            illnesses = sum(Illnesses),
            hospitalizations = sum(Hospitalizations),
            fatalities = sum(Fatalities),
            status=sum(),
            .groups = "keep") %>%
  arrange(desc(illnesses)) -> df.num

#Histogram of year and illnesses
ggplot(df.num)+geom_col(mapping=aes(x=Year, y=illnesses), fill="maroon")
```